# (PART) Chapter II {.unnumbered}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, width = 300, warning = FALSE, out.width = "50%", fig.show = 'hold', fig.align = "center")
library(here)
library(tinytex)
library(ggplot2)
library(dplyr)
library(tidyr)
library(rlang)
library(usethis)
library(bookdown)
library(knitr)
library(rmarkdown)
library(spelling)
```

# **Animal Architects: A Conceptual Model for Geomorphic and Pedological Animal Activity**

**Very Tentative "Abstract"**

The subject of this review is: how do animals impact the physical evolution of soils and terrestrial landscapes? There is an abundance of literature on the roles of individual animals, in both case studies and reviews. I first propose a framework or conceptual model for classifying these behaviors. The model incorporates both lateral and vertical dimensions of animal activity. Vertically, animals either mix or layer soil. Laterally, animals create variety in surface structures and/or more broadly influence erosion processes. It also addresses the dual roles of animals in creating heterogeneity and homogeneity in terrestrial landscapes and soils. I next use the conceptual model to evaluate the emphases, successes, and limitations of current numerical modeling approaches. Advanced measurement techniques like optically stimulated luminescence, fallout radionuclide tracers, or image analysis, are increasingly applied to bioturbation research. These methods are often used to generate model parameters that provide generalized descriptions of bioturbation processes. There are several approaches to this model construction covered here, each of which with a unique emphasis and limitations.

## Introduction

Soil is the most complex system in the natural world. At a fundamental level, it encompasses biological, chemical, and physical sciences. At an earth-system level, it governs the planet’s atmosphere, ecosystems, and water cycle. And at a disciplinary level, it unites agronomists, ecologists, geomorphologists, and hydrologists. With so much breadth, our knowledge of soils is vast—which is fitting for the "most complex and unparsimonious of all natural science entities” (Johnson et al., 2005). However, this diversity of work, to be impact and organized, requires a central guiding principle. To understand where we are going, we need to look at where we've been.

Pedology, the study of soil formation, was first defined by Friedrich Albert Fallou in his 1862 book “Pedology or General and Special Soil Science” (Fallou, 1862). Since then, pedology has been defined and refined by many authors, all circling Hans Jenny's basic definition, “the section of the domain of soil science that studies the soil body in its natural position” (Jenny, 1941). It explains that soil is predicable; that its formation is governed by factors and processes that can be quantified. However, although pedology is so clearly defined, its singular subject remains hard to characterize, as the same soil profile, as is it today, could be the focus of manys -ists: agronomists, ecologists, soil geomorphologists, hydrologists, etc. Beyond that, soils change through time, so there are countless profiles with countless properties through countless time.

The role of a pedologist is to simplify this immense amount of information it into a narrative, and any narrative needs structure. Pedologists, like all scientists, rely on simplified, conceptual models to understand their system.

The most influential, although not the first (citation), is Jenny's clorpt model, where soil is defined as a function of climate, organisms, relief, parent material, and time (Jenny, 1941). The cloprt model provides a framework for thinking about soils and situates soils in their broader environment. It also allows for connections to be drawn between hydrology (climate), geomorphology (relief), geology (parent material), and ecology (organisms).

There are two lessons we can learn from this: 1) conceptual models are invaluable to understanding the natural world, and 2) animals are important.

Animals have long been recognized to play an important role in the history of our soils and landscapes. Charles Darwin was the first to recognize the influence of animals, earthworms, in the development of soil morphology (Darwin, 1881), and although organisms were once underrepresented in pedological discourse (Johnson & Schaetzl, 2015), their prevalence has substantially increased in recent years.

Soil or soil-adjacent scientists use many terms to describe the role living organisms play in a soil system. Broadly pedoturbation, bioturbation, fauna- and floraturbation, or biomixing (Wilkenson et al., 2009), or more specifically mounding, cratering, or burrowing (Johnson et al., 2005a), to name a few. For this discussion, I use a generalized term: animal activity. Animal activity has the advantage of being implicitly process-based, i.e. animal activities are distinct from the evidence of animal activities. It also encourages thinking about the animals themselves and their relationships to each other and their environment. And finally, it suggests a degree of unintentionality. Animals, oftentimes including humans, don't understand soils or soil science. All the evidences we study are the unintentional consequences of their lifestyles, often over generations. Thus, the product of animal activities are a refecltion of their ecological characters rather.

There exist already a number of excellent reviews on the various roles and categorizations’ of animal activity in soils (Bétard, 2021; Gabet et al., 2003; Johnson et al., 2005b, 2005a; Johnson & Watson-Stegner, 1987; Wilkinson et al., 2009). So, although there is overlap, this paper's emphasis is on the quantification of these categories and how they can be integrated in a holistic, qualitative framework for describing soil fauna activities. The overarching goal of this paper is to take a first, albeit small, step towards a standardized understanding of how and why animal activity is measured that can be applied across the discipline of soil science... To that end, I begin with a discussion of the types of relevant animal behaviors and how they can be categorized.

## Conceptual models

**Animal behaviors and classifications**

Animals are extremely diverse in their individual behaviors, and species ecology drives their role on the landscape. Ants, termites, gophers, and earthworms, among others, each occupy unique roles, niches, that translate to unique changes in soils and landscapes. Here, I consider any behavior that translocates soil, either vertically or laterally, across a terrestrial environment. There exists substantial inter-species difference as well–exemplified in the ecological classifications of earthworms discussed later. Throughout this cross- and inter-species variability, there are a two trends–or axes–that arise. One, animals, introduce diversity on lateral and vertical scales. And two, animals create unique impacts across timescales (Figure 1).

```{r, echo = FALSE}
include_graphics((here("./_figures/Four panel animals.svg")), dpi= NULL)
```
