# Erosion Pin Data Workflow

```{r setup, include = FALSE}
# chunk options
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, width = 300, warning = FALSE, out.width = "45%", fig.show = 'hold', fig.align = "center")

# libraries needed
libs <- c("lidR", "shapefiles", "sf", "terra", "raster", "tidyr", "tidyverse","dplyr", "ggplot2", "easypackages", "spatialEco","here", "performance", "RColorBrewer", "lme4", "nlme")

# install missing libraries
installed_libs <- libs %in% rownames(installed.packages())
if (any(installed_libs == F)) {
  install.packages(libs[!installed_libs])
}

# load libraries
lapply(libs, library, character.only = T)

# enter the file path for the highest level folder you're working in 
data_folder <- "C:/Users/natha/Box/box_data/_data/_erosion_pins/"

# when a file is needed, call the hert() function
# for example: data_frame = read.csv(hert("more_data/measurements_data.csv"))
hert <- function(file) {
  file_path = paste(data_folder, file, sep = "")
  return(file_path)
}


```

## Measurement Data

Pull in the data from box and do some pre-processing.

```{r}

# Pull in the data from box and some pre-processing.
input <- read.csv(hert("measurements_data.csv"))

# we need to first average the duplicate measurements for each pin.
input$mm_ch <- (input$mm1_ch + input$mm2_ch) / 2 # for changes
input$mm_bl <- (input$mm2_bl + input$mm2_bl.1) / 2 # for baseline
  
# convert date to days past
input$date_date <- as.Date(input$date) # format date column
input$dayof <- as.numeric(input$date_date) - 20283 + 195 # convert to days since 1970-01-01
  
# create and index column, a unique ID for each mm
input$index <- (paste(input$site, input$forest, input$transect, input$slope_pos, input$pin.., sep = "_"))
input$index_num <- as.numeric(factor(input$index))

# filter the dataset to remove unneeded columns
clean <- input %>%
  select(date_date, dayof, index_num, worms, site, 
         forest, transect, slope_pos, pin.., mm_ch, mm_bl) %>% 
  rename(pin = pin..)
  

# arrange pins unique pin, in order of date
clean_sorted <- clean %>% 
  arrange(index_num) %>% # sort
  mutate(dt = 0) %>% # add a dt column for later
  mutate(dmm = 0) # add a dmm column for later

# number of unique pins, should be 216
nindex <- length(unique(clean_sorted$index))

# create a list, where each item is a data frame with all the measurements of each pin
df_list <- vector(mode = "list", length = nindex) # empty list
for(i in 1:nindex) {
  df_list[[i]] <- clean_sorted %>% filter(index_num == i)
} 


```

Differencing erosion pin measurments

```{r}

for(i in 1:nindex){ # out for loop selects each data frame (individual pin) in the list
  for(j in 2:6) { # inner for loop goes throughe each column of list 
    df_list[[i]][j,12] <- df_list[[i]][j, 2] - df_list[[i]][j - 1, 2] # subtracts today's day of from last measurement's
    df_list[[i]][j,13] <- df_list[[i]][j - 1, 11] - df_list[[i]][j, 10] # subtracts today's mm from last baseline
  }
}
  

#  rebuild the data frame
  master_df <- df_list[[1]] # jank start to the for loop, needs a base
for(i in 2:nindex) { # for loop to stack dataframes one list at a time
  master_df <- merge(master_df, df_list[[i]], all = TRUE)
} 

print(master_df)
  
```

Statistical analysis

```{r}
# clean up data frame to remove outliers (greater than 3sd from the mean)
master <- master_df %>%
  mutate(abs_dmm = abs(dmm)) %>% # create abs_dmm column
  filter(abs_dmm < sd(master_df$dmm)*3 & dt > 0) # filter to remove inital sites and outliers

```


```{r}
  
#'###################### [for individual pin data]

masteri <- master %>% 
  group_by(forest, transect, slope_pos, pin) %>%
  mutate(dhdt = abs_dmm / dt) %>% 
  summarize(dhdt_ = sum(dhdt), abs_dh = sum(abs_dmm), worms = first(worms)) %>% 
  mutate(dhdt = dhdt_ * 365 / 10 ) # dhdt in cm/yr


# arrange the data for later plotting
  masteri$forest <- factor(masteri$forest, 
                           levels=c("ASH", "LRE", "LRW", "MAG", "WD", "LRJ"))

  
  
# FOREST SPECEFIC boxplot for dhdt
ggplot(data = masteri, mapping = aes(x = forest, y = dhdt, fill = worms)) +
  geom_boxplot() +
  facet_wrap(~slope_pos) +
  scale_fill_manual(values = c("JW" = "#D6604D", "EW"= "#4393C3")) +
  scale_y_continuous(name = "Change in Pins (cm / yr)") +
  scale_x_discrete(name = "Forest") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(fill = "Worms")

# WORM boxplot for abs_dmm/dt
ggplot(data = masteri, mapping = aes(x = worms, y = dhdt, fill = worms)) +
  geom_boxplot() +
  facet_wrap(~slope_pos)+ 
  scale_fill_manual(values = c("JW" = "#D6604D", "EW"= "#4393C3")) +
  scale_fill_manual(values = c("JW" = "#D6604D", "EW"= "#4393C3")) +
  scale_y_continuous(name = "Change in Pins (cm / yr)") +
  scale_x_discrete(name = "Worms") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none")

# FOREST SPECEFIC boxplot for dhdt
ggplot(data = masteri, mapping = aes(x = forest, y = dhdt, fill = worms)) +
  geom_boxplot() +
  facet_wrap(~slope_pos) +
  scale_fill_manual(values = c("JW" = "#D6604D", "EW"= "#4393C3")) +
  scale_y_continuous(name = "Change in Pins (cm / yr)") +
  scale_x_discrete(name = "Forest") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(fill = "Worms")

###############################non -abs
masteri <- master %>% 
  group_by(forest, transect, slope_pos, pin) %>%
  mutate(dhdt = abs_dmm / dt) %>% 
  summarize(dhdt_ = sum(dhdt), abs_dh = sum(abs_dmm), worms = first(worms)) %>% 
  mutate(dhdt = dhdt_ * 365 / 10 ) # dhdt in cm/yr


# arrange the data for later plotting
  masteri$forest <- factor(masteri$forest, 
                           levels=c("ASH", "LRE", "LRW", "MAG", "WD", "LRJ"))

  
  
# FOREST SPECEFIC boxplot for dhdt
ggplot(data = masteri, mapping = aes(x = forest, y = dhdt, fill = worms)) +
  geom_boxplot() +
  facet_wrap(~slope_pos) +
  scale_fill_manual(values = c("JW" = "#D6604D", "EW"= "#4393C3")) +
  scale_y_continuous(name = "Change in Pins (cm / yr)") +
  scale_x_discrete(name = "Forest") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(fill = "Worms")

# WORM boxplot for abs_dmm/dt
ggplot(data = masteri, mapping = aes(x = worms, y = dhdt, fill = worms)) +
  geom_boxplot() +
  facet_wrap(~slope_pos)+ 
  scale_fill_manual(values = c("JW" = "#D6604D", "EW"= "#4393C3")) +
  scale_fill_manual(values = c("JW" = "#D6604D", "EW"= "#4393C3")) +
  scale_y_continuous(name = "Change in Pins (cm / yr)") +
  scale_x_discrete(name = "Worms") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none")

# FOREST SPECEFIC boxplot for dhdt
ggplot(data = masteri, mapping = aes(x = forest, y = dhdt, fill = worms)) +
  geom_boxplot() +
  facet_wrap(~slope_pos) +
  scale_fill_manual(values = c("JW" = "#D6604D", "EW"= "#4393C3")) +
  scale_y_continuous(name = "Change in Pins (cm / yr)") +
  scale_x_discrete(name = "Forest") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(fill = "Worms")

```

## GPS Data

Import GPS data points saved in the G-drive (collected using Field Maps and downloaded through Arc Online). Transform into NAD83 zone 15N.

```{r}
# load erosion pin points,
arb_gps_raw <- read.csv(hert("_transect_gps/Erosion_Pin_Transects_1.csv")) %>% 
  select(Latitude, Longitude) %>% 
  filter(Latitude < 45) # arb and lr are in the same file, split into 2

lr_gps_raw <- read.csv("C:/Users/natha/Documents/local_data/_data/_erosion_pins/_transect_gps/Erosion_Pin_Transects_1.csv") %>% 
  select(Latitude, Longitude) %>% 
  filter(Latitude > 45)

# above points are theoretically in WGS84. this tells r they are (which in R the crs is EPSG: 4326). this also makes them a simple features feature (sf)
arb_gps <- st_as_sf(arb_gps_raw, coords = c("Longitude", "Latitude"), crs = 4326)
lr_gps <- st_as_sf(lr_gps_raw, coords = c("Longitude", "Latitude"), crs = 4326)

# transform from WGS84 to NAD83 zone 15N to match lidar (EPSG: 4269)
arb_NAD <- st_transform(arb_gps, crs = 26915)
lr_NAD <- st_transform(lr_gps, crs = 26915)

lr_NAD <- lr_NAD[-31,] # remove junk point
```

Import dems from drive (created in 26_mapping).

```{r}
# lr
lr_1m_dem <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/lr_1m_dem")
lr_1m_slope <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/lr_1m_slope")
lr_1m_curvature <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/lr_1m_curvature")

# arb
arb_1m_dem <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_arb_topo/arb_1m_dem")
arb_1m_slope <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_arb_topo/arb_1m_slope")
arb_1m_curvature <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_arb_topo/arb_1m_curvature")
```

Extract elevation, slope, and curvature (in 1/m) at gps points.

```{r}
# arb
arb_NAD$elevation <- terra::extract(arb_1m_dem, arb_NAD)[,2]
arb_NAD$slope <- terra::extract(arb_1m_slope, arb_NAD)[,2]
arb_NAD$curvature <- terra::extract(arb_1m_curvature, arb_NAD)[,2]

# lr
lr_NAD$elevation <- terra::extract(lr_1m_dem, lr_NAD)[,2]
lr_NAD$slope <- terra::extract(lr_1m_slope, lr_NAD)[,2]
lr_NAD$curvature <- terra::extract(lr_1m_curvature, lr_NAD)[,2]
```

Export data sets to drive.

```{r}
# export as shape files (.shp)
write_sf(arb_NAD, "C:/Users/natha/Documents/local_data/_data/_erosion_pins/_arb_geo_data_shp/arb_geo_data.shp")
write_sf(lr_NAD, "C:/Users/natha/Documents/local_data/_data/_erosion_pins/_lr_geo_data_shp/lr_geo_data.shp")

# export as tables (.csv)
write.csv(arb_NAD, "C:/Users/natha/Documents/local_data/_data/_erosion_pins/arb_geo_data.csv", quote = 1)
write.csv(lr_NAD, "C:/Users/natha/Documents/local_data/_data/_erosion_pins/lr_geo_data.csv", quote = 1)
```

## Visualizing etc.

### Spatial Plots

Import GPS and dem data.

```{r, eval = TRUE}
# import shape files
arb_geo_shp <- read_sf("C:/Users/natha/Documents/local_data/_data/_erosion_pins/_arb_geo_data_shp/arb_geo_data.shp")
lr_geo_shp <- read_sf("C:/Users/natha/Documents/local_data/_data/_erosion_pins/_lr_geo_data_shp/lr_geo_data.shp")

# import elevation slope curvature data
arb_geo_data <- read.csv("C:/Users/natha/Documents/local_data/_data/_erosion_pins/arb_geo_data.csv")
lr_geo_data <- read.csv("C:/Users/natha/Documents/local_data/_data/_erosion_pins/lr_geo_data.csv")

# merge two above
arb_geo_data[1:2] <- st_coordinates(arb_geo_shp)
arb_geo_data <- rename(arb_geo_data, "latitude" = X, "longitude" = geometry)

lr_geo_data[1:2] <- st_coordinates(lr_geo_shp)
lr_geo_data <- rename(lr_geo_data, "latitude" = X, "longitude" = geometry)

# dems
lr_1m_dem <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/lr_1m_dem")
lr_1m_slope <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/lr_1m_slope")
lr_1m_curvature <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/lr_1m_curvature")

arb_1m_dem <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_arb_topo/arb_1m_dem")
arb_1m_slope <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_arb_topo/arb_1m_slope")
arb_1m_curvature <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_arb_topo/arb_1m_curvature")
```

Plotting fun

```{r, eval = TRUE}
arb_geo_data$site = NA
arb_geo_data[1:15,]$site = "ASH"
arb_geo_data[16:30,]$site = "MAG"
arb_geo_data[31:45,]$site = "WD"

ggplot(arb_geo_data, mapping = aes(latitude, longitude, group = site, color = curvature)) +
  geom_point(size = 2.5) +
  facet_wrap(~site)

```
