# Fallout Radionulcides

```{r setup, include = FALSE}
# chunk options
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, width = 300, warning = FALSE, out.width = "45%", fig.show = 'hold', fig.align = "center")

# libraries needed
libs <- c("lidR", "shapefiles", "sf", "terra", "raster", "tidyr", "dplyr", "ggplot2", "easypackages", "spatialEco")

# install missing libraries
installed_libs <- libs %in% rownames(installed.packages())
if (any(installed_libs == F)) {
  install.packages(libs[!installed_libs])
}

# load libraries
lapply(libs, library, character.only = T)
```

## Dataset Building

### Extracting elevation, slope, and curvature from DEMS to GPS points.

Import GPS data points saved in the G-drive (collected using Field Maps and downloaded through Arc Online). Transform into NAD83 zone 15N.

```{r}
# load frn transect points
arb_gps_raw <- 
  read.csv("C:/Users/natha/Documents/local_data/_data/_fallout_radionuclides/_transect_gps/FRN_Sampling_Points_0.csv") %>% 
  select(Latitude, Longitude) %>% 
  filter(Latitude < 45) # arb and lr are in the same file, split into 2
  
lr_gps_raw <- 
  read.csv("C:/Users/natha/Documents/local_data/_data/_fallout_radionuclides/_transect_gps/FRN_Sampling_Points_0.csv") %>% 
  select(Latitude, Longitude) %>% 
  filter(Latitude > 45)

# above points are theoretically in WGS84. this tells r they are (which in R the crs is EPSG: 4326). this also makes them a simple features feature (sf)
arb_gps <- st_as_sf(arb_gps_raw, coords = c("Longitude", "Latitude"), crs = 4326)
lr_gps <- st_as_sf(lr_gps_raw, coords = c("Longitude", "Latitude"), crs = 4326)

# transform from WGS84 to NAD83 zone 15N to match lidar (EPSG: 4269)
arb_NAD <- st_transform(arb_gps, crs = 26915)
lr_NAD <- st_transform(lr_gps, crs = 26915)
```

Import dems from drive (created in 26_mapping).

```{r}
# lr
lr_1m_dem <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/lr_1m_dem")
lr_1m_slope <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/lr_1m_slope")
lr_1m_curvature <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/lr_1m_curvature")

# arb
arb_1m_dem <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_arb_topo/arb_1m_dem")
arb_1m_slope <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_arb_topo/arb_1m_slope")
arb_1m_curvature <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_arb_topo/arb_1m_curvature")
```

Extract elevation, slope, and curvature (in 1/m) at gps points.

```{r}
# arb
arb_NAD$elevation <- terra::extract(arb_1m_dem, arb_NAD)[,2]
arb_NAD$slope <- terra::extract(arb_1m_slope, arb_NAD)[,2]
arb_NAD$curvature <- terra::extract(arb_1m_curvature, arb_NAD)[,2]

# lr
lr_NAD$elevation <- terra::extract(lr_1m_dem, lr_NAD)[,2]
lr_NAD$slope <- terra::extract(lr_1m_slope, lr_NAD)[,2]
lr_NAD$curvature <- terra::extract(lr_1m_curvature, lr_NAD)[,2]
```

Export data sets to drive.

```{r}
# export as shape files (.shp)
write_sf(arb_NAD, "C:/Users/natha/Documents/local_data/_data/_fallout_radionuclides/_arb_geo_data_shp/arb_geo_data.shp")
write_sf(lr_NAD, "C:/Users/natha/Documents/local_data/_data/_fallout_radionuclides/_lr_geo_data_shp/lr_geo_data.shp")

# export as tables (.csv)
write.csv(arb_NAD, "C:/Users/natha/Documents/local_data/_data/_fallout_radionuclides/arb_geo_data.csv", quote = 1)
write.csv(lr_NAD, "C:/Users/natha/Documents/local_data/_data/_fallout_radionuclides/lr_geo_data.csv", quote = 1)
```

