# Mapping

```{r setup, include = FALSE}
# chunk options
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, width = 300, warning = FALSE, out.width = "45%", fig.show = 'hold', fig.align = "center")

# libraries needed
libs <- c("lidR", "shapefiles", "sf", "terra", "raster", "tidyr", "dplyr", "ggplot2", "easypackages", "spatialEco", "rasterVis")

# install missing libraries
installed_libs <- libs %in% rownames(installed.packages())
if (any(installed_libs == F)) {
  install.packages(libs[!installed_libs])
}

# load libraries
lapply(libs, library, character.only = T)
```

Most data in this project is collected across the Minnesota Landscape Arboretum and Lake Rebecca Park Reserve, so mapping is focoused there.

## Creating DEMs (and other rasters) from MN LiDAR data

Technical resources

-   sf manual: <https://cran.r-project.org/web/packages/sf/sf.pdf>

-   lidR manual: <https://r-lidar.github.io/lidRbook/>

-   shapefiles manual: <https://cran.r-project.org/web/packages/shapefiles/shapefiles.pdf>

-   MNtopo <https://files.dnr.state.mn.us/aboutdnr/gis/mntopo/mntopo_help_document.pdf>

-   EPSG info: <https://www.nceas.ucsb.edu/sites/default/files/2020-04/OverviewCoordinateReferenceSystems.pdf>

-   SF: https://r-spatial.github.io/sf/articles/sf1.html#reading-and-writing

Elevation, slope, and curvature can be extracted from Minnesota LiDAR data found at MNTopo (<http://arcgis.dnr.state.mn.us/maps/mntopo/>). The raw LiDAR data can be downloaded directly from the website. LiDAR data is in NAD83 w/ elevation in meters.

```{r}
# load laz fliles from arb (in NAD83). filter for only ground-classified points (-keep_class 2)
arb_las1 <- readLAS("C:/Users/natha/Documents/local_data/_data/_mapping/_arb_topo/_lidar/4342-05-13.laz", filter = "-keep_class 2")
arb_las2 <- readLAS("C:/Users/natha/Documents/local_data/_data/_mapping/_arb_topo/_lidar/4342-05-12.laz", filter = "-keep_class 2")

# load laz files from lr
lr_las1 <- readLAS("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/_lidar/3542-30-08.laz", filter = "-keep_class 2")
lr_las2 <- readLAS("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/_lidar/3542-30-09.laz", filter = "-keep_class 2")
lr_las3 <- readLAS("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/_lidar/3542-31-08.laz", filter = "-keep_class 2")
lr_las4 <- readLAS("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/_lidar/3542-31-09.laz", filter = "-keep_class 2")

```

Transform coordinate reference system to NAD83 zone 15N (ESPG 26915). Fit a digital terrain model to the data using the triangular irregular network algorithm (not as computationally intense, default settings). Terrain function (<https://www.rdocumentation.org/packages/raster/versions/3.6-32/topics/terrain>)

```{r}
# transform to NAD83 zone 15N (ESPG 26915)
epsg(arb_las1) # report the current coordinate system
st_crs(arb_las1) <- 26915 # set to  NAD83 zone 15N
st_crs(arb_las2) <- 26915
st_crs(lr_las1) <- 26915
st_crs(lr_las2) <- 26915
st_crs(lr_las3) <- 26915
st_crs(lr_las4) <- 26915

# fit DEM using triangular irregular network, this is similar to LAStools stuff too
arb_dem1 <- rasterize_terrain(arb_las1, res = 1, algorithm = tin())
arb_dem2 <- rasterize_terrain(arb_las2, res = 1, algorithm = tin())

lr_dem1 <- rasterize_terrain(lr_las1, res = 1, algorithm = tin())
lr_dem2 <- rasterize_terrain(lr_las2, res = 1, algorithm = tin())
lr_dem3 <- rasterize_terrain(lr_las3, res = 1, algorithm = tin())
lr_dem4 <- rasterize_terrain(lr_las4, res = 1, algorithm = tin())


# merge DEMS
arb_dem <- mosaic(arb_dem1, arb_dem2)

lr_dem <- mosaic(lr_dem1, lr_dem2, lr_dem3, lr_dem4)
```

Create new rasters using dems for slope and curvature. Slope is in degrees. Curvature is calculated in the direction of maximum slope (profile). Positive values indicate upward covex.

```{r}
# slope (in degrees)
arb_slope <- terrain(arb_dem, "slope", unit = "degrees")
lr_slope <- terrain(lr_dem, "slope", unit = "degrees")

# curvature
arb_curvature <- curvature(arb_dem, type = "profile")
lr_curvature <- curvature(lr_dem, type = "profile")
```

Export models to G-Drive.

```{r}
# arb
terra::writeRaster(arb_dem, filetype = "GTiff", filename = file.path("C:/Users/natha/Documents/local_data/_data/_mapping/_arb_topo/arb_1m_dem"), overwrite=TRUE)

terra::writeRaster(arb_slope, filetype = "GTiff", filename = file.path("C:/Users/natha/Documents/local_data/_data/_mapping/_arb_topo/arb_1m_slope"), overwrite=TRUE)

terra::writeRaster(arb_curvature, filetype = "GTiff", filename = file.path("C:/Users/natha/Documents/local_data/_data/_mapping/_arb_topo/arb_1m_curvature"), overwrite=TRUE)

# lr
terra::writeRaster(lr_dem, filetype = "GTiff", filename = file.path("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/lr_1m_dem"), overwrite=TRUE)

terra::writeRaster(lr_slope, filetype = "GTiff", filename = file.path("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/lr_1m_slope"), overwrite=TRUE)

terra::writeRaster(lr_curvature, filetype = "GTiff", filename = file.path("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/lr_1m_curvature"), overwrite=TRUE)
```

## Visualzing Sites

```{r, eval = TRUE}
# import data
lr_1m_dem <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/lr_1m_dem")
lr_1m_slope <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/lr_1m_slope")
lr_1m_curvature <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_lake_rebecca_topo/lr_1m_curvature")

arb_1m_dem <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_arb_topo/arb_1m_dem")
arb_1m_slope <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_arb_topo/arb_1m_slope")
arb_1m_curvature <- rast("C:/Users/natha/Documents/local_data/_data/_mapping/_arb_topo/arb_1m_curvature")
```


```{r, eval = TRUE, cache = TRUE}

# Crop the raster
# cropped_raster <- crop(arb_1m_dem, extent(c(min(arb_geo_data$latitude) + 500,
#                                             max(arb_geo_data$latitude) + 500,
#                                             min(arb_geo_data$longitude) + 500,
#                                             max(arb_geo_data$longitude + 500))))

# format dem
arb_dem_map <- as.data.frame(arb_1m_dem, xy = TRUE) %>% 
  rename("elevation" = "Z")

ggplot(data = arb_dem_map) +
  geom_tile(mapping = aes(x = x, y = y, fill = elevation)) +
  scale_fill_viridis_c() +
  labs(title = "Digital Elevation Model", fill = "Elevation") +
  coord_equal() +
  theme_minimal()

```




