## Models and Code

```{r setup, echo = FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
library(tinytex)
library(ggplot2)
library(dplyr)
library(tidyr)
library(rlang)
library(usethis)
library(bookdown)
library(knitr)
library(rmarkdown)
library(reticulate)
use_condaenv("C:/Users/natha/anaconda3/envs/animalarchitects_main")
```

```{python p_setup, echo = FALSE, include=FALSE}
# import python packages
import matplotlib.pyplot as plt
import numpy as np
# %matplotlib inline
```

*This first code chunk defines the shared structures and variables used by the following code chunks. It has no outputs itself.*

```{python model_setup, echo = TRUE}

# build structures for data inputs. These do not get overwritten.
dz = 0.1 # m
z = np.arange(0, 3, dz, dtype=float) # create an array of depths
y = np.zeros(z.shape, dtype=float) # create a blank array for y, copies shape of z
p = np.zeros(z.shape, dtype=float)# create a blank array for p, copies shape of z

# [set] initial p values
p += 1 # mass soil / volume soil

# [set] total time
total_time = 10000 # years

# [set] erosion/denudation rate. erosion (-)
T = 0 # m / yr

# Define the exponential diffusion equation
def D(z, D0, zb):
    return D0 * np.exp(-z/zb) # in m2/yr

# adjust timestep size
dt = 0.2 * dz**2 / 10**(-3) # sets time step size, based on Courant–Friedrichs–Lewy condition. Currently a general, low value
time_steps = int(total_time / dt) # calculates total number of time steps


def OSL_plt_defaults():
  plt.ylim(-1.2, 0)
  plt.legend()
  plt.xlabel('Age since exposure to surface (yr)')
  plt.ylabel('Depth (m)')
  plt.show()

```

*This code plots the differing exponential D(z) equations.*

```{python diffusion_dc}
# diffusion depth comparisons (_dc0)

# Calculate y values, one array per plot
D_dc1 = D(z, 10**(-5), 0.50) # Kirkby
D_dc2 = D(z, 9.81 * 10**(-5), 0.28) # Johnson
D_dc3 = D(z, 1.95*10**(-5), 0.28) # Roman-Sanchez

# Plot the equation
plt.clf()
plt.plot(D_dc1, -z, label = "Kirkby")
plt.plot(D_dc2, -z, label = "Johnson")
plt.plot(D_dc3, -z, label = "Roman-Sanchez")
plt.xlabel('diffusion (m2/yr)')
plt.ylabel('depth (m)')
plt.grid(True)
plt.legend()
plt.show()

```

*This code uses the above the equations to simulate an age profile of soil grains created by biodiffusion, local mixing, only.*

```{python local_biodiffusion1, echo=TRUE}

# define OSL diffusion only function
def OSL_diffusion(z, y, p, dz, dt, T, D):
  y_copy = y.copy()
  for i in range(time_steps):
    qy = -1 * D[0:-1] * np.diff(y_copy * p) / dz
    diff = -1 * np.diff(qy) / dz
    advec = T * -np.diff(y_copy * p) / dz
    dydt = diff + advec[0:-1] + 1 
    y_copy[1:-1] += dydt * dt
  return y_copy

Johnson = OSL_diffusion(z, y, p, dz, dt, T, D(z, 9.81 * 10**(-5), 0.28))
Kirkby = OSL_diffusion(z, y, p, dz, dt, T, D(z, 10**(-5), 0.50))


plt.clf()
plt.plot(Johnson, -z, label = "Johnson et al. (2014)")
plt.plot(Kirkby, -z, label = "Kirkby (1985)")
OSL_plt_defaults()

```

*This code generates a visualization of the impact of erosion on a theoretical tracer profile. 1) tracer profile at steady state, 2) tracer profile moves downwards under deposition, and 3) tracer profile moves upwards under erosion. Soil surface at z = 0.*

```{r erosion_visulization, warning=FALSE}
eq = function(z){
  0.01 * exp(-z/0.15/100)
  }

tdf <- data.frame(z = rep((1:100), 3), ts = c(rep(1, times=100), rep(2, times=100), rep(3, times=100)), y = 0)

tdf$y[1:100] = eq(tdf$z[1:100])
tdf$y[101:200] = eq(tdf$z[1:100] - 20)
tdf$y[201:300] = eq(tdf$z[1:100] + 20)
tdf$y[101:120] = NA

ggplot(data = tdf, mapping = aes(y = z, x = y, group = ts)) +
  scale_y_reverse(name = "z (cm)") +
  scale_x_reverse(name = "y", 
                  breaks = c()) +
  #coord_flip() +
  geom_line() +
  ggtitle("Visualization of net-zero, deposistion, and erosion.") +
  facet_wrap(~ts, scales = "fixed") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

```

*This code adds advection, by erosion and deposition, to the model.*

```{python local_biodiffusion_advection, echo = TRUE}

T1 = OSL_diffusion(z, y, p, dz, dt, 0, D(z, 9.81 * 10**(-5), 0.28))
T2 = OSL_diffusion(z, y, p, dz, dt, -0.00005, D(z, 9.81 * 10**(-5), 0.28))
T3 = OSL_diffusion(z, y, p, dz, dt, -0.0001, D(z, 9.81 * 10**(-5), 0.28))
T4 = OSL_diffusion(z, y, p, dz, dt, 0.00005, D(z, 9.81 * 10**(-5), 0.28))
T5 = OSL_diffusion(z, y, p, dz, dt, 0.00001, D(z, 9.81 * 10**(-5), 0.28))

plt.clf()
plt.plot(T1, -z, label = "T = 0", color = "Black")
plt.plot(T2, -z, label = "T = -0.00005", linestyle='--')
plt.plot(T3, -z, label = "T = -0.0001", linestyle='--')
plt.plot(T4, -z, label = "T = 0.00005", linestyle='--')
plt.plot(T5, -z, label = "T = 0.0001", linestyle='--')
OSL_plt_defaults()


```

*This code uses the above function to do some "tracer" plotting with different deposition values.*

```{python nonOSL_diffusion, echo = TRUE}

# define OSL diffusion, with a tracer, function
def local_diffusion(z, y, p, dz, dt, T, D):
  y_copy = y.copy()
  for i in range(time_steps):
    qy = -1 * D[0:-1] * np.diff(y_copy * p) / dz
    diff = -1 * np.diff(qy) / dz
    advec = T * -np.diff(y_copy * p) / dz
    dydt = diff + advec[0:-1]
    y_copy[1:-1] += dydt * dt
  return y_copy

smart0 = local_diffusion(z, y, p, dz, dt, 0, D(z, 9.81 * 10**(-5), 0.28))
smart = local_diffusion(z, y, p, dz, dt, 0.00001, D(z, 9.81 * 10**(-5), 0.28))
smart2 = local_diffusion(z, y, p, dz, dt, 0.0001, D(z, 9.81 * 10**(-5), 0.28))
smart3 = local_diffusion(z, y, p, dz, dt, 0.001, D(z, 9.81 * 10**(-5), 0.28))

plt.clf()
plt.plot(y, -z, label = "original", color = "Black", linestyle='--')
plt.plot(smart0, -z, label = "T = 0", color = "Black")
plt.plot(smart, -z, label = "T = +0.00001", linestyle='--')
plt.plot(smart2, -z, label = "T = +0.0001", linestyle='--')
plt.plot(smart3, -z, label = "T = +0.001", linestyle='--')
OSL_plt_defaults()
```

*This code does the above equation*

```{python local_isotope, echo = TRUE}

# [set] initial y values
yi = y.copy()
yi[z == 0] += 1 # mass or activity y / mass soil

# [set] initial p values
pi = p.copy()
pi += 1 # mass soil / volume soil


# define isotope local diffusion function, constant D
def local_diffusion(z, yi, p, dz, dt, T, D):
  y_copy = y.copy()
  for i in range(time_steps):
    qy = -1 * D[0:-1] * np.diff(y_copy * p) / dz
    diff = -1 * np.diff(qy) / dz
    advec = T * -np.diff(y_copy * p) / dz
    dydt = diff + advec[0:-1]
    y_copy[1:-1] += dydt * dt
  return y_copy

```
